Component response rate variation underlies the stability of highly complex finite systems

### 2020-03-23 notebook/ms.tex
Insert two additional references that were recommended by reviewer 3

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies the stability of complex systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Component response rate variation underlies the stability of complex
systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

In 1972, May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} first
demonstrated that randomly assembled systems of sufficient complexity
are almost inevitably unstable given infinitesimally small
perturbations. Complexity in this case is defined by the size of the
system (i.e., the number of potentially interacting components; \(S\)),
its connectance (i.e., the probability that one component will interact
with another; \(C\)), and the variance of interaction strengths
(\(\sigma^{2}\))\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
May's finding that the probability of local stability falls to near zero
given a sufficiently high threshold of \(\sigma\sqrt{SC}\) is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{5}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{6},\protect\hyperlink{ref-Gray2009}{7}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{8},\protect\hyperlink{ref-MacArthur2010}{9}},
and
socio-economic\textsuperscript{\protect\hyperlink{ref-May2008}{10}--\protect\hyperlink{ref-Bardoscia2017}{13}}
networks. As such, identifying general principles that affect stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
\(-1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis on
\(\mathbf{M}\), with the system being stable if the real parts of all
eigenvalues (\(\lambda\)), and therefore the leading eigenvalue
(\(\lambda_{max}\)), are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -d\) (\(-d\) is the mean value of diagonal
elements) and \(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and \(\sigma\)
increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < d\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) (\(\sigma^{2}_{\gamma}\)) might be expected to
decrease system stability by introducing a new source of variation into
the system and thereby increasing \(\sigma\). Here I show that, despite
higher \(\sigma\), realistic complex systems (in which \(S\) is high but
finite) are actually more likely to be stable if their individual
component response rates vary. My results are robust across commonly
observed network structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks.

@@

\textbf{Component response rates of random complex systems}. Complex
systems (\(\mathbf{M}\)) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\boldsymbol{\gamma}\)). Both \(\mathbf{A}\) and
\(\boldsymbol{\gamma}\) are square \(S \times S\) matrices. Rows in
\(\mathbf{A}\) define how a given component \(i\) is affected by each
component \(j\) in the system, including itself (where \(i = j\)).
Off-diagonal elements of \(\mathbf{A}\) are independent and identically
distributed (i.i.d), and diagonal elements are set to \(A_{ii} = -1\) as
in May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\boldsymbol{\gamma}\) are positive, and off-diagonal
elements are set to zero (i.e. \(\boldsymbol{\gamma}\) is a diagonal
matrix with positive support). The distribution of
\(diag(\boldsymbol{\gamma})\) over \(S\) components thereby models the
distribution of component response rates. The dynamics of the entire
system \(\mathbf{M}\) can be defined as
follows\textsuperscript{\protect\hyperlink{ref-Patel2018}{19}},

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\boldsymbol{\gamma}\) is set to the
identity matrix (diagonal elements all equal 1) and
\(\mathbf{M} = \mathbf{A}\). Under these conditions, eigenvalues of
\(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} in a
circle centred at \((-d, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

\textbf{Effect of \(\mathbf{\sigma^{2}_{\gamma}}\) on \(\mathbf{M}\)
(co)variation}. The value of \(\Re(\lambda_{max})\), and therefore
system stability, can be estimated from five properties of
\(\mathbf{M}\)\textsuperscript{\protect\hyperlink{ref-Tang2014b}{20}}.
These properties include (1) system size (\(S\)), (2) mean
self-regulation of components (\(d\)), (3) mean interaction strength
between components (\(\mu\)), (4) the variance of between component
interaction strengths (hereafter \(\sigma^{2}_{M}\), to distinguish from
\(\sigma^{2}_{A}\) and \(\sigma^{2}_{\gamma}\)), and (5) the correlation
of interaction strengths between components, \(M_{ij}\) and \(M_{ji}\)
(\(\rho\)). Positive \(\sigma^{2}_{\gamma}\) does not change \(S\), nor
does it necessarily change \(E[d]\) or \(E[\mu]\). What
\(\sigma^{2}_{\gamma}\) does change is the total variation in component
interaction strengths (\(\sigma^{2}_{M}\)), and \(\rho\). Introducing
variation in \(\gamma\) increases the total variation in the system.
Variation in the off-diagonal elements of \(\mathbf{M}\) is described by
the joint variation of two random variables,

@@

The increase in \(\sigma^{2}_{M}\) caused by \(\sigma^{2}_\gamma\) can
be visualised from the eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \boldsymbol{\gamma}\textbf{A}\) (Fig. 1). Given \(d = 0\)
and \(C = 1\), the distribution of eigenvalues of \(\textbf{A}\) and
\(\textbf{M}\) lie within a circle of a radius \(\sigma_{A}\sqrt{S}\)
and \(\sigma_{M}\sqrt{S}\), respectively (Fig. 1a vs.~1b). If
\(d \neq 0\), positive \(\sigma^{2}_\gamma\) changes the distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{21}--\protect\hyperlink{ref-Stone2017}{23}},
potentially affecting stability (Fig. 1c vs.~1d).

Given \(\sigma^{2}_\gamma = 0\), \(\Re(\lambda_{max})\) decreases
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{24}},

@@

I also simulated 100000 \(\mathbf{M = \gamma A}\) for three types of
random networks that are typically interpreted as modelling three types
of interspecific ecological
interactions\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{25}}.
These interaction types are competitive, mutualist, and predator-prey,
as modelled by off-diagonal elements that are constrained to be
negative, positive, or paired such that if \(A_{ij} > 0\) then
\(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{A}\)). As
\(S\) increased, a higher number of stable \(\mathbf{M}\) relative to
\(\mathbf{A}\) was observed for competitor and predator-prey, but not
mutualist, systems. A higher number of stable systems was observed
whenever \(S > 12\) and \(S > 40\) for competitive and predator-prey
systems, respectively (note that \(\rho < 0\) for predator-prey systems,
making stability more likely overall). The stability of mutualist
systems was never affected by \(\sigma^{2}_{\gamma}\).

@@

\textbf{Simulation of structured \(\mathbf{M}\) across \(\mathbf{S}\)}.
To investigate how \(\sigma^{2}_{\gamma}\) affects the stability of
commonly observed network structures, I simulated one million
\(\mathbf{M = \gamma A}\) for
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}, and
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{17},\protect\hyperlink{ref-Williams2000}{18}}
networks. In all of these networks, rules determining the presence or
absence of an interaction between components \(i\) and \(j\) constrain
the overall structure of the network. In small-world networks,
interactions between components are constrained so that the expected
degree of separation between any two components increases in proportion
to \(\log(S)\)\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}}.
In scale-free networks, the distribution of the number of components
with which a focal component interacts follows a power law; a few
components have many interactions while most components have few
interactions\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}}. In
cascade food webs, species are ranked and interactions are constrained
such that a species \(i\) can only feed on \(j\) if the rank of
\(i > j\).

@@

\textbf{System feasibility given \(\mathbf{\sigma^{2}_{\gamma}}\)} For
complex systems in which individual system components represent the
density of some tangible quantity, it is relevant to consider the
feasibility of the system. Feasibilility assumes that values of all
components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{5},\protect\hyperlink{ref-Dougoud2018}{26},\protect\hyperlink{ref-Song2018}{27}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{26}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices tested for stability. I emphasise that
\(\gamma\) is not interpreted as population density in this analysis,
but instead as a fundamental property of species life history such as
expected generation time. Feasibility was unaffected by
\(\sigma^{2}_{\gamma}\) and instead occurred with a fixed probability of
\(1/2^{S}\), consistent with a recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{28}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

\textbf{Targeted manipulation of \(\boldsymbol{\gamma}\)}. To further
investigate the potential of \(\sigma^{2}_{\gamma}\) to be stabilising,
I used a genetic algorithm. Genetic algorithms are heuristic tools that
mimic evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{29}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\Re(\lambda_{max})\) demonstrated the potential for
\(\sigma^{2}_{\gamma}\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (see Supplementary
Information), meaning that stability could often be achieved by
manipulating \(S\) \(\gamma\) values rather than \(S \times S\)
\(\mathbf{M}\) elements (i.e., by manipulating component response rates
rather than interactions between components).

@@

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{30}},

@@

\hypertarget{ref-Mougi2012}{}
3. Mougi, A. \& Kondoh, M. Diversity of interaction types and ecological
community stability. \emph{Science} \textbf{337,} 349--351 (2012).

\hypertarget{ref-Allesina2015}{}
4. Allesina, S. \emph{et al.} Predicting the stability of large
structured food webs. \emph{Nature Communications} \textbf{6,} 7842
(2015).

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{1}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{unicode=true,
            pdftitle={Component response rate variation underlies the stability of highly complex finite systems},
            pdfauthor={A. Bradley Duthie ( alexander.duthie@stir.ac.uk )},
            colorlinks=true,
            linkcolor=blue,
            citecolor=Blue,
            urlcolor=Blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Component response rate variation underlies the stability of highly
complex finite systems}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{A. Bradley Duthie (
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{Biological and Environmental Sciences, University of Stirling, Stirling,
UK, FK9 4LA}

@@

In 1972, May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} first
demonstrated that randomly assembled systems of sufficient complexity
are almost inevitably unstable given infinitesimally small
perturbations. Complexity in this case is defined by the size of the
system (i.e., the number of potentially interacting components; \(S\)),
its connectance (i.e., the probability that one component will interact
with another; \(C\)), and the variance of interaction strengths
(\(\sigma^{2}\))\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
May's finding that the probability of local stability falls to near zero
given a sufficiently high threshold of \(\sigma\sqrt{SC}\) is broadly
relevant for understanding the dynamics and persistence of systems such
as
ecological\textsuperscript{\protect\hyperlink{ref-May1972}{1}--\protect\hyperlink{ref-Grilli2017}{6}},
neurological\textsuperscript{\protect\hyperlink{ref-Gray2008}{7},\protect\hyperlink{ref-Gray2009}{8}},
biochemical\textsuperscript{\protect\hyperlink{ref-Rosenfeld2009}{9},\protect\hyperlink{ref-MacArthur2010}{10}},
and
socio-economic\textsuperscript{\protect\hyperlink{ref-May2008}{11}--\protect\hyperlink{ref-Bardoscia2017}{14}}
networks. As such, identifying general principles that affect stability
in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{12}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
\(-1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}.
Local system stability is assessed using eigenanalysis on
\(\mathbf{M}\), with the system being stable if the real parts of all
eigenvalues (\(\lambda\)), and therefore the leading eigenvalue
(\(\lambda_{max}\)), are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} within a
circle centred at \(\Re = -d\) (\(-d\) is the mean value of diagonal
elements) and \(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{5}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and \(\sigma\)
increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < d\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) (\(\sigma^{2}_{\gamma}\)) might be expected to
decrease system stability by introducing a new source of variation into
the system and thereby increasing \(\sigma\). Here I show that, despite
higher \(\sigma\), realistic complex systems (in which \(S\) is high but
finite) are actually more likely to be stable if their individual
component response rates vary. My results are robust across commonly
observed network structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{18},\protect\hyperlink{ref-Williams2000}{19}}
networks.

@@

\textbf{Component response rates of random complex systems}. Complex
systems (\(\mathbf{M}\)) are built from two matrices, one modelling
component interactions (\(\mathbf{A}\)), and second modelling component
response rates (\(\boldsymbol{\gamma}\)). Both \(\mathbf{A}\) and
\(\boldsymbol{\gamma}\) are square \(S \times S\) matrices. Rows in
\(\mathbf{A}\) define how a given component \(i\) is affected by each
component \(j\) in the system, including itself (where \(i = j\)).
Off-diagonal elements of \(\mathbf{A}\) are independent and identically
distributed (i.i.d), and diagonal elements are set to \(A_{ii} = -1\) as
in May\textsuperscript{\protect\hyperlink{ref-May1972}{1}}. Diagonal
elements of \(\boldsymbol{\gamma}\) are positive, and off-diagonal
elements are set to zero (i.e. \(\boldsymbol{\gamma}\) is a diagonal
matrix with positive support). The distribution of
\(diag(\boldsymbol{\gamma})\) over \(S\) components thereby models the
distribution of component response rates. The dynamics of the entire
system \(\mathbf{M}\) can be defined as
follows\textsuperscript{\protect\hyperlink{ref-Patel2018}{20}},

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma^{2}_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma^{2}_{\gamma} = 0\)), \(\boldsymbol{\gamma}\) is set to the
identity matrix (diagonal elements all equal 1) and
\(\mathbf{M} = \mathbf{A}\). Under these conditions, eigenvalues of
\(\mathbf{M}\) are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{15}} in a
circle centred at \((-d, 0)\) with a radius of
\(\sigma \sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
(Fig. 1a).

\textbf{Effect of \(\mathbf{\sigma^{2}_{\gamma}}\) on \(\mathbf{M}\)
(co)variation}. The value of \(\Re(\lambda_{max})\), and therefore
system stability, can be estimated from five properties of
\(\mathbf{M}\)\textsuperscript{\protect\hyperlink{ref-Tang2014b}{21}}.
These properties include (1) system size (\(S\)), (2) mean
self-regulation of components (\(d\)), (3) mean interaction strength
between components (\(\mu\)), (4) the variance of between component
interaction strengths (hereafter \(\sigma^{2}_{M}\), to distinguish from
\(\sigma^{2}_{A}\) and \(\sigma^{2}_{\gamma}\)), and (5) the correlation
of interaction strengths between components, \(M_{ij}\) and \(M_{ji}\)
(\(\rho\))\textsuperscript{\protect\hyperlink{ref-Sommers1988}{22}}.
Positive \(\sigma^{2}_{\gamma}\) does not change \(S\), nor does it
necessarily change \(E[d]\) or \(E[\mu]\). What \(\sigma^{2}_{\gamma}\)
does change is the total variation in component interaction strengths
(\(\sigma^{2}_{M}\)), and \(\rho\). Introducing variation in \(\gamma\)
increases the total variation in the system. Variation in the
off-diagonal elements of \(\mathbf{M}\) is described by the joint
variation of two random variables,

@@

The increase in \(\sigma^{2}_{M}\) caused by \(\sigma^{2}_\gamma\) can
be visualised from the eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \boldsymbol{\gamma}\textbf{A}\) (Fig. 1). Given \(d = 0\)
and \(C = 1\), the distribution of eigenvalues of \(\textbf{A}\) and
\(\textbf{M}\) lie within a circle of a radius \(\sigma_{A}\sqrt{S}\)
and \(\sigma_{M}\sqrt{S}\), respectively (Fig. 1a vs.~1b). If
\(d \neq 0\), positive \(\sigma^{2}_\gamma\) changes the distribution of
eigenvalues\textsuperscript{\protect\hyperlink{ref-Ahmadian2015}{23}--\protect\hyperlink{ref-Stone2017}{25}},
potentially affecting stability (Fig. 1c vs.~1d).

Given \(\sigma^{2}_\gamma = 0\), \(\Re(\lambda_{max})\) decreases
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{26}},

@@

I also simulated 100000 \(\mathbf{M = \gamma A}\) for three types of
random networks that are typically interpreted as modelling three types
of interspecific ecological
interactions\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2011}{27}}.
These interaction types are competitive, mutualist, and predator-prey,
as modelled by off-diagonal elements that are constrained to be
negative, positive, or paired such that if \(A_{ij} > 0\) then
\(A_{ji} < 0\),
respectively\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
(but are otherwise identical to the purely random \(\mathbf{A}\)). As
\(S\) increased, a higher number of stable \(\mathbf{M}\) relative to
\(\mathbf{A}\) was observed for competitor and predator-prey, but not
mutualist, systems. A higher number of stable systems was observed
whenever \(S > 12\) and \(S > 40\) for competitive and predator-prey
systems, respectively (note that \(\rho < 0\) for predator-prey systems,
making stability more likely overall). The stability of mutualist
systems was never affected by \(\sigma^{2}_{\gamma}\).

@@

\textbf{Simulation of structured \(\mathbf{M}\) across \(\mathbf{S}\)}.
To investigate how \(\sigma^{2}_{\gamma}\) affects the stability of
commonly observed network structures, I simulated one million
\(\mathbf{M = \gamma A}\) for
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}, and
cascade food
web\textsuperscript{\protect\hyperlink{ref-Solow1998}{18},\protect\hyperlink{ref-Williams2000}{19}}
networks. In all of these networks, rules determining the presence or
absence of an interaction between components \(i\) and \(j\) constrain
the overall structure of the network. In small-world networks,
interactions between components are constrained so that the expected
degree of separation between any two components increases in proportion
to \(\log(S)\)\textsuperscript{\protect\hyperlink{ref-Watts1998}{16}}.
In scale-free networks, the distribution of the number of components
with which a focal component interacts follows a power law; a few
components have many interactions while most components have few
interactions\textsuperscript{\protect\hyperlink{ref-Albert2002}{17}}. In
cascade food webs, species are ranked and interactions are constrained
such that a species \(i\) can only feed on \(j\) if the rank of
\(i > j\).

@@

\textbf{System feasibility given \(\mathbf{\sigma^{2}_{\gamma}}\)} For
complex systems in which individual system components represent the
density of some tangible quantity, it is relevant to consider the
feasibility of the system. Feasibilility assumes that values of all
components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{6},\protect\hyperlink{ref-Dougoud2018}{28},\protect\hyperlink{ref-Song2018}{29}}.
This is of particular interest for ecological communities because
population density (\(N\)) cannot take negative values, meaning that
ecological systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{28}}.
While my results are intended to be general to all complex systems, and
not restricted to species networks, I have also performed a feasibility
analysis on all matrices tested for stability. I emphasise that
\(\gamma\) is not interpreted as population density in this analysis,
but instead as a fundamental property of species life history such as
expected generation time. Feasibility was unaffected by
\(\sigma^{2}_{\gamma}\) and instead occurred with a fixed probability of
\(1/2^{S}\), consistent with a recent proof by Serván et
al.\textsuperscript{\protect\hyperlink{ref-Servan2018}{30}} (see
Supplementary Information). Hence, for pure interacting species
networks, variation in component response rate (i.e., species generation
time) does not affect stability at biologically realistic species
densities.

\textbf{Targeted manipulation of \(\boldsymbol{\gamma}\)}. To further
investigate the potential of \(\sigma^{2}_{\gamma}\) to be stabilising,
I used a genetic algorithm. Genetic algorithms are heuristic tools that
mimic evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{31}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\Re(\lambda_{max})\) demonstrated the potential for
\(\sigma^{2}_{\gamma}\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (see Supplementary
Information), meaning that stability could often be achieved by
manipulating \(S\) \(\gamma\) values rather than \(S \times S\)
\(\mathbf{M}\) elements (i.e., by manipulating component response rates
rather than interactions between components).

@@

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{32}},

@@

\hypertarget{ref-Townsend2010a}{}
3. Townsend, S. E., Haydon, D. T. \& Matthews, L. On the generality of
stability-complexity relationships in Lotka-Volterra ecosystems.
\emph{Journal of Theoretical Biology} \textbf{267,} 243--251 (2010).
